[Re] Modeling GABA Alterations in Schizophrenia: A Link Between Impaired Inhibition and Gamma and Beta Auditory Entrainment

### 2017-06-20 article/Metzner-2017.tex
fixed issues aaronshifman

In the original article the authors conduct an MEG study of auditory
entrainment and find changes in gamma and beta range entrainment found
in schizophrenic patients. These changes in oscillatory dynamics in
patients are important for two reasons: First, gamma range oscillations
emerge in a variety of different tasks and are thought to underlie many
cognitive processes (see e.g. \autocite{Fries2005,Fries2015}). Furthermore, impaired gamma
range oscillations in schizophrenic patientas have been found in most of
these tasks and might offer an explanation for the cognitive deficits
found in patients. Therefore, the study of the mechanisms underlying
oscillatory entrainment might shed light on the basic principles that
are responsible for a wide range of deficits in schizophrenic patients
Second, oscillatory entrainment is a promising candidate for a
neurophysiological biomarker for schizophrenia
\autocite{Siekmeier2015}. In the original article they develop two
models that are able to account for the changes they find in their
experimental data. One is a biophysically detailed network model and the
other one a simplified network model. In both models a prolonged decay
at GABAergic inhibitory synapses causes the reduction in gamma
entrainment and the increase in beta entrainment. The simplified network
model, despite its many simplifications, offers insight into the
mechanisms underlying these entrainment deficits while focusing on a few
key parameters. This makes it easy to distill and understand underlying
dynamics and mechanisms and, at the same time, makes simulations
computationally very inexpensive allowing for an extensive exploration
of the parameter space. Additionally, again because of the simplicity,
the model can easily be extended to study the effect other parameters
(e.g.~sparsity of connections, different populations of inhibitory
neurons, different types of synaptic receptors,\ldots{}), without
loosing much of its simplicity.

@@

The model was implemented solely from the paper description, since the
original code is not publicly available. The model is a simple model
consisting of two neural populations (excitatory and inhibitory cells).
Individual cells are modeled as theta neurons (for a detailed discussion
of this neuron model see \autocite{Boergers2003}). The neuron is
described by a single variable \(\theta\), which can be regarded as the
membrane voltage, subject to the following dynamics

@@

where \(\alpha _j=\pm 1\) controls excitation and inhibition, \(g_{jk}\)
is the synaptic strength from cell \(j\) to cell \(k\) and \(s_{jk}\) is
the synaptic gating variable from cell \(j\) to cell \(k\). Synaptic
gating variables are subject to the following dynamics

@@

Each population connects to itself and to the other with an all-to-all
connectivity. Both populations also have two sources of input, the
oscillatory drive input and a background noise input. The drive input
periodically sends spikes to both populations with a given frequency,
whereas the background noise input sends noise spikes at times drawn
from a Poisson distribution. Table \ref{tbl:summary} summarizes the
network model. Tables \ref{tbl:modparameters} and
\ref{tbl:simparameters} list the parameters of the model and the
simulations, their definitions and values, respectively.

@@

All differential equations are solved using a simple forward Euler
scheme. As in the original article a single simulation simulates a 500ms
trial and the time step is chosen such that this results in 2\^{}13 =
8192 data points. However, the main results are unaffected by a smaller
time step.

@@

After having looked at the model output averaged over 20 trials with
different noise, we also show single trial data which exemplify the main
model features , as was done in the original article. Figures
\ref{fig:Vierling6} and fig:Vierling7 show the model output in
response to 40 Hz drive input for the control and the schizophrenia
network, respectively. The strong entrainment in the control case, the
reduction of entrainment and the emergence of a subharmonic 20 Hz
component are again clearly visible. However, as before, in our model
implementation the emergent 20 Hz component is less pronounced than in
the original implementation (best seen in the excitatory population
activitydisplayed in the raster plot of \ref{fig:Vierling7}).

>>>>

In the original article the authors conduct an MEG study of auditory
entrainment and find changes in gamma and beta range entrainment found
in schizophrenic patients. These changes in oscillatory dynamics in
patients are important for two reasons: first, gamma range oscillations
emerge in a variety of different tasks and are thought to underlie many
cognitive processes (see e.g. \autocite{Fries2005,Fries2015}). Furthermore, impaired gamma
range oscillations in schizophrenic patientas have been found in most of
these tasks and might offer an explanation for the cognitive deficits
found in patients. Therefore, the study of the mechanisms underlying
oscillatory entrainment might shed light on the basic principles that
are responsible for a wide range of deficits in schizophrenic patients.
Second, oscillatory entrainment is a promising candidate for a
neurophysiological biomarker for schizophrenia
\autocite{Siekmeier2015}. In the original article they develop two
models that are able to account for the changes they find in their
experimental data. One is a biophysically detailed network model and the
other one a simplified network model. In both models a prolonged decay
at GABAergic inhibitory synapses causes the reduction in gamma
entrainment and the increase in beta entrainment. The simplified network
model, despite its many simplifications, offers insight into the
mechanisms underlying these entrainment deficits while focusing on a few
key parameters. This makes it easy to distill and understand underlying
dynamics and mechanisms and, at the same time, makes simulations
computationally very inexpensive allowing for an extensive exploration
of the parameter space. Additionally, again because of the simplicity,
the model can easily be extended to study the effect other parameters
(e.g.~sparsity of connections, different populations of inhibitory
neurons, different types of synaptic receptors,\ldots{}), without
loosing much of its simplicity.

@@

The model was implemented solely from the paper description, since the
original code is not publicly available. The model is a simple model
consisting of two neural populations (excitatory and inhibitory cells).
Individual cells are modeled as theta neurons (for a detailed discussion
of this neuron model see \autocite{Boergers2003}). The neuron is
described by a single variable \(\theta\), which can be regarded as the
neuron state, subject to the following dynamics

@@

where \(\alpha _j\) controls excitation and inhibition, i.e. is $+1$ for excitatory synapses and $-1$ for inhibitory ones,
\(g_{jk}\)
is the synaptic strength from cell \(j\) to cell \(k\) and \(s_{jk}\) is
the synaptic gating variable from cell \(j\) to cell \(k\). Synaptic
gating variables are subject to the following dynamics

@@

Each population connects to itself and to the other with an all-to-all
connectivity. Both populations also have two sources of input, the
oscillatory drive input and a background noise input. The drive input
periodically sends spikes to both populations with a given frequency. In order to generate the spikes
a drive cell is implemented (also a theta neuron), which receives an applied current so that its spike frequency
matches the driving frequency. This drive cell is then connected to all cells in the network.
The background noise input sends noise spikes at times drawn
from a Poisson distribution. Table \ref{tbl:summary} summarizes the
network model. Tables \ref{tbl:modparameters} and
\ref{tbl:simparameters} list the parameters of the model and the
simulations, their definitions and values, respectively.

@@

All differential equations were solved using a simple forward Euler
scheme. As in the original article a single simulation simulated a 500ms
trial and the time step was chosen such that this results in 2\^{}13 =
8192 data points. However, the main results are unaffected by a smaller
time step.

@@

After having looked at the model output averaged over 20 trials with
different noise, we also show single trial data which exemplify the main
model features , as was done in the original article. Figures
\ref{fig:Vierling6} and \ref{fig:Vierling7} show the model output in
response to 40 Hz drive input for the control and the schizophrenia
network, respectively. The strong entrainment in the control case, the
reduction of entrainment and the emergence of a subharmonic 20 Hz
component are again clearly visible. However, as before, in our model
implementation the emergent 20 Hz component is less pronounced than in
the original implementation (best seen in the excitatory population
activitydisplayed in the raster plot of \ref{fig:Vierling7}).
